Asssesment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-09-29 analysis/paper/paper.Rmd
Arregla comentarios introducción de Leandro

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a; @fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020]. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

#LEAN: Antes de empezar este parráfo, me parece que deberías describir las maneras en que se suele calcular el SAM en la literatura (algo así creo que habías incluido para la AGU).

However, computed as the leading Empirical Orthogonal Function (EOF) of Sea Level Pressure or low-level geopotential height, the SAM spatial structure contains noticeable deviations from this zonally symmetric description, particularly in the Pacific Ocean region. These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009]. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in renalysis products prior to the sattelite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].

#LEAN: También me parece que faltaría agregar dos parráfos más, uno relacionado con las tendencias del SAM y otro relacionado con los impactos conocidos del SAM.

#LEAN: Otra cosa que no veo y que me parece que sería importante decir es sobre la estructura vertical del SAM (sería bueno también agregar la referencia de Baldwin que es la que inspira un poco el estudio).

Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability by constructing two indices which aim to capture exclusively the variability of the symmetric and asymmetric component each. These indices are constructed for each vertical level, so that vertical coherence could be assessed. We analyse their temporal variability and trends.

#acá puede ser confuso porque cuando hablas de "cada indice" a veces te referis a los indices de todos los niveles verticales, en otro de solo uno que describa la estratósfera y otro la tropósfera y en otro solo al de la troposfera.

We then study the spatial patterns described by the variability exclusive to each index. Finally, we investigate their relationship with temperature and precipitation anomalies.

#LEAN: Faltaría acá después describir las secciones del trabajo

@@

# Methods

@@

To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and precipitation. We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree lagitude by 5\degree longitude. For precipitation, we use monthly, 0.5\degree latitude by 0.5\degree longitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions. In such case, using naive p-values to test for significance leads to misleading results [@walker1914;@katz1991]. While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue. Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.

@@

# Results

## Temporal evolution

@@

```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}
annual_trend <- indices %>%
  copy() %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(lev, type = term)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  .[, season := "Annual"]

indices %>% 
  copy() %>% 
  # .[time > as.Date("1990-01-01")] %>%
  .[, season := season(time)] %>% 
  .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time = time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL] %>% 
  rbind(annual_trend) %>% 
  .[, estimate := estimate*3600*24*365*10] %>% 
  .[, std.error := std.error*3600*24*365*10] %>% 
  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
  setnames("type", "term") %>% 
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

@@

## Spatial patterns

@@

## Impacts

@@

The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile. This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region. On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North. This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009, @zamboni2010]. This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].

@@

The two-year periodcty we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation. There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980, @watson2014, @zhang2020], so it's not unlikely that the SAM would be similarly connected. However establishing this link would require further research.

>>>>

The Southern Annular Mode (SAM) is the main mode of variability in the Southern Hemisphere extratropical circulation [@rogers1982] in daily, monthly, and decadal timescales [@baldwin2001a;@fogt2006] and exerts an important influence in weather conditions such as temperature and precipitation anomalies and sea ice concentration [@fogt2020]. Its positive phase is traditionally described as anomalously low pressures over Antarctica surrounded by a ring of anomalous high pressures in middle-to-high latitudes.

Most authors describe the SAM as a zonally symmetric pattern, a fact that is reflected not only in its name, but also in the various methods used to characterise it. Of the several different indices presented in the literature, many of them are based on zonal means of sea level pressure or geopotential height [@ho2012]. @gong1999 defined the SAM index as the zonal mean sea level pressure differene between 40\degree S and 60\degree S, which is also the definition used by the station-based index in @marshall2003. @baldwin2009 proposed defining the Northern and Annular modes as the leading EOF of the zonally averaged geopotential height at each level.

Even though these indices are based on zonal averages, the spatial the spatial structure of the SAM computed from them contains noticeable deviations from zonal symmetry, particularly in the Pacific Ocean region. These zonal asymmetries have not been widely studied, but previous work suggest that they strongly modulate the regional impacts of the SAM, going as far as reversing its relationship between precipitation in South America [@silvestri2009]. At the very least, the fact that the SAM is not entirely zonally symmetric hinders our ability to reconstruct its historical variability prior to the availabilty of dense observations in the Southern Hemisphere [@jones2009].

At least some of the variabilty associated with the zonal asymmetries of the SAM is probably forced by the tropics. In particular, ENSO-like variability affects the Southern Hemisphere extratopics through the Pacific-South American Pattern [@mo1987; @kidson1988;@karoly1989], whose wave train proyects strongly onto the zonal anomalies corresponding to the SAM in the Pacific sector. And although the relationship between ENSO and SAM is far from simple, tropical influences on the SAM have been observed [@fan2007; @fogt2011; @clem2013].

Positive trends in SAM index have been documented by various researchers using different indices mostly on boreal Summer and Autumn [@fogt2020 and references therein]. These trends are thought of driven primarily by stratospheric ozone depletion and understood in the context of zonal mean variables. However, it's not clear how or if the asymmetric component responds to this forcing, or whether its variabilty could be masking influencing the observed trends independendly.

Similarly unclear are the specific impacts of the zonally asymmetric component of the SAM. Positive phase of the SAM is associated with generally colder temperatures over Antarctica and warmer temperatures at higher latitudes [@jones2019] (and viceversa for negative SAM), but there are significant deviations from this zonal mean response, notably in the Antarctic Peninsula and the South Atlantic [@fogt2012]. The SAM signal in precipitation behaves similarly, although with even greater deviation from zonal symmetry [@lim2016]. The importance of zonal asymmetries of the SAM in these impacts have been studied in certain regions. For example, the SAM-precipitation relationship in Southeastern South America and Southern Brazil can be explained by the PSA-like zonally asymetric circulation associated with the SAM [@silvestry2009; @rosso2018]. @fan2007 also found that precipitation in East Asia was impacted by the variabilty of only the Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal variability of the asymmetric component of the SAM with the exception of @fogt2012. However, their methods based on composites of positive and negative SAM events leads to some issues, such as spatial patterns derived from as little as 4 cases and from imbalanced periods (for example, 5 of the 7 cases in their DJF SAM+ composite are from later than 1988, whereas all of the 8 years in their DJF SAM- composite are from earlier than 1988). This is particularly important due to the inhomogeneities in renalysis products prior to the satellite era and the possible change in the asymmetric structure of the SAM [@silvestri2009].

Moreover, @fogt2012 studied the zonal asymmetric component of the SAM only in sea level pressure. Zonal asymmetries in the SAM spatial pattern are fairly barotropic throughout the troposphere, but they change dramatically in the stratosphere [@baldwin2009].

Our objective is, then, to systematically characterise the zonally asymmetric and symmetric components of the SAM variability. For each level, we construct two indices which aim to capture exclusively the variability of the symmetric and asymmetric component respectively. We assess their vertical structure and coherence, temporal variability and trends. We then study the spatial patterns described by the variability exclusive to each index. Finally, we investigate their relationship with temperature and precipitation anomalies.

In the Section \ref{sec:methods} we describe the methods. In Section \ref{sec:results}.\ref{sec:temporal} we describe the temporal variability and vertical coherence of the inidices. In  Section \ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of geopotential height associated with them. In Section \ref{sec:results}.\ref{sec:impacts}, we study their relationship with surface-level tempreature and precipitation.

@@

# Methods 
\label{sec:methods}

@@

To describe the Southern Annular Mode and its variability we used monthly geopotential height at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@hersbach2020] for the period 1979 to 2018. We restrict our analysis to the post-satellite era to avoid any confounding factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and precipitation. We use temperature data from NOAA's Merged Land Ocean Global Surface Temperature Analysis V4.0.1 [@vose2012; @smith2008], which blends land and ocean temperature analysis into a monthly global grid 5\degree\ longitude by 5\degree\ latitude For precipitation, we use monthly, 0.5\degree\ longitude by 0.5\degree\ latitude data from the Global Precipitation Climatology Centre [@schneider2015].

@@

At 2.5\degree by 2.5\degree resolution, a single regression field is composed of thousands of regressions. In such case, using naive p-values to test for significance leads to misleading results [@walker1914; @katz1991]. While there are multiple proposed solutions in the literature, @wilks2016 suggests that adjunting p-values by controlling for the False Discovery Rate [@benjamini1995] is a simple and effective method to ameliorate this issue. Therefore, p-values showed in regression fields are all adjusted following @benjamini1995.

@@

# Results
\label{sec:results}

## Temporal evolution
\label{sec:temporal}

@@

```{r trends, fig.cap = "Decadal trends at each level for annual (row 1) and seasonal values (rows 2 to 5) for the period 1979-2018 and for the (column a) Full SAM index, (column b) Asymmetric SAM index, and (column c) Symmetric SAM index. Shading indicates the 95\\% confidence interval.", fig.width = 6, fig.height=4, fig.env="figure*"}

indices %>% 
  copy() %>% 
  .[, season := season(time)] %>% 
  rbind(., copy(.)[, season := "Annual"]) %>%
    .[, estimate_norm := estimate_norm/sd(estimate_norm), by = .(lev, term)] %>%
  .[, FitLm(estimate_norm, time, se = TRUE), by = .(type = term, lev, season)] %>% 
  rm_intercept() %>% 
  .[, term := NULL]  %>%
  .[, estimate := estimate*3600*24*365*10] %>%
  .[, std.error := std.error*3600*24*365*10] %>%
  # .[, estimate := estimate/(3600*24*365*10)] %>%
  # .[, std.error := std.error/(3600*24*365*10)] %>%
  .[, pval := Pvaluate(estimate, std.error, df, "fdr")] %>% 
  setnames("type", "term") %>%
  .[, season := forcats::fct_relevel(season, "Annual")] %>% 
  ggplot(aes(lev, estimate)) +
  geom_hline(yintercept = 0) +
  geom_ribbon(aes(ymax = estimate + std.error*1.96, ymin = estimate - std.error*1.96), 
              fill = "#95a3ab", color = "black", 
              alpha = .6, size = 0.1) +
  geom_line() +
  # stat_subset(aes(subset = pval <= 0.05), shape = 4, size = 2) +
  scale_fill_brewer(palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_level() +
  scale_y_continuous("Standard deviations per decade") +
  coord_flip() +
  facet_grid(season ~ term,labeller = labeller(term = lab_sam)) +
  tag_facets("cr") 
```

@@

## Spatial patterns
\label{sec:spatial}

@@

## Impacts
\label{sec:impacts}

@@

The anomalous circulation at 700 hPa associated with the Symmetric SAM (panel c.1) indicate anomalous Easterly flow over Southern Chile. This leads to reduced influx of moist air from the Pacific Ocean which, is the main source of precipitable water in that region. On the other hand, the anomalous circulation associated with positive values of Asymmetric SAM (panel b.1) in the Atlantic is anticyclonic in the South and cyclonic in the North. This creates anomalous South-Easterly flow over Southeastern South America, which inhibits the flow of the Low Level Jet to the region [@silvestri2009; @zamboni2010]. This same pattern was found to be associated with increased precipitation in Southern Brazil during South Atlantic Convergence Zone events [@rosso2018].

@@

The two-year periodcty we found in the stratospheric Symmetric SAM might point to a link between the SAM and the Quasi Biennial Oscillation. There is evidence of influence between the QBO and the Northern Annular Mode [e.g. @holton1980; @watson2014; @zhang2020], so it's not unlikely that the SAM would be similarly connected. However establishing this link would require further research.

### 2020-09-29 analysis/paper/paper.tex
Arregla comentarios introducción de Leandro

\title{New method to describe the zonal symmetries and asymmetries of the
Southern Annular Mode}

@@

The Southern Annular Mode (SAM) is the main mode of variability in the
Southern Hemisphere extratropical circulation \citep{rogers1982} in
daily, monthly, and decadal timescales {[}\citet{baldwin2001a};
fogt2006{]} and exerts an important influence in weather conditions such
as temperature and precipitation anomalies and sea ice concentration
\citep{fogt2020}. Its positive phase is traditionally described as
anomalously low pressures over Antarctica surrounded by a ring of
anomalous high pressures in middle-to-high latitudes.

However, computed as the leading Empirical Orthogonal Function (EOF) of
Sea Level Pressure or low-level geopotential height, the SAM spatial
structure contains noticeable deviations from this zonally symmetric
description, particularly in the Pacific Ocean region. These zonal
asymmetries are not widely studied, but previous work suggest that they
strongly modulate the regional impacts of the SAM, going as far as
reversing its relationship between precipitation in South America
\citep{silvestri2009}. At the very least, the fact that the SAM is not
entirely zonally symmetric hinders our ability to reconstruct its
historical variability prior to the availabilty of dense observations in
the Southern Hemisphere \citep{jones2009}.

We are not aware of any previous work which quantifies the temporal
variability of the asymmetric component of the SAM with the exception of
\citet{fogt2012}. However, their methods based on composites of positive
and negative SAM events leads to some issues, such as spatial patterns
derived from as little as 4 cases and from imbalanced periods (for
example, 5 of the 7 cases in their DJF SAM+ composite are from later
than 1988, whereas all of the 8 years in their DJF SAM- composite are
from earlier than 1988). This is particularly important due to the
inhomogeneities in renalysis products prior to the sattelite era and the
possible change in the asymmetric structure of the SAM
\citep{silvestri2009}.

Our objective is, then, to systematically characterise the zonally
asymmetric component of the SAM variability by constructing two indices
which aim to capture exclusively the variability of the symmetric and
asymmetric component each. We then analyse their temporal variability,
trends and vertical coherence. We study the spatial patterns described
by the variability exclusive to each index. Finally, we investigate
theri relationship with temperature and recipitation anomalies.

@@

@@

To describe the Southern Annular Mode and its variability we used
monthly geopotential height at 2.5\degree longitude by
2.5\degree latitude of horizontal resolution and 37 vertical isobaric
levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018. We
restrict our analysis to the post-satellite era to avoid any confounding
factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and
precipitation. We use temperature data from NOAA's Merged Land Ocean
Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
which blends land and ocean temperature analysis into a monthly global
grid 5\degree lagitude by 5\degree longitude. For precipitation, we use
monthly, 0.5\degree latitude by 0.5\degree longitude data from the
Global Precipitation Climatology Centre \citep{schneider2015}.

@@

@@

@@

@@

@@

The anomalous circulation at 700 hPa associated with the Symmetric SAM
(panel c.1) indicate anomalous Easterly flow over Southern Chile. This
leads to reduced influx of moist air from the Pacific Ocean which, is
the main source of precipitable water in that region. On the other hand,
the anomalous circulation associated with positive values of Asymmetric
SAM (panel b.1) in the Atlantic is anticyclonic in the South and
cyclonic in the North. This creates anomalous South-Easterly flow over
Southeastern South America, which inhibits the flow of the Low Level Jet
to the region \citep[\citet{zamboni2010}]{silvestri2009}. This same
pattern was found to be associated with increased precipitation in
Southern Brazil during South Atlantic Convergence Zone events
\citep{rosso2018}.

@@

The two-year periodcty we found in the stratospheric Symmetric SAM might
point to a link between the SAM and the Quasi Biennial Oscillation.
There is evidence of influence between the QBO and the Northern Annular
Mode \citep[e.g.][\citet{watson2014}, \citet{zhang2020}]{holton1980}, so
it's not unlikely that the SAM would be similarly connected. However
establishing this link would require further research.

>>>>

\title{Asssesment of zonal symmetric and asymmetric components of the Southern
Annular Mode using a novel approach}

@@

The Southern Annular Mode (SAM) is the main mode of variability in the
Southern Hemisphere extratropical circulation \citep{rogers1982} in
daily, monthly, and decadal timescales \citep{baldwin2001a, fogt2006}
and exerts an important influence in weather conditions such as
temperature and precipitation anomalies and sea ice concentration
\citep{fogt2020}. Its positive phase is traditionally described as
anomalously low pressures over Antarctica surrounded by a ring of
anomalous high pressures in middle-to-high latitudes.

Most authors describe the SAM as a zonally symmetric pattern, a fact
that is reflected not only in its name, but also in the various methods
used to characterise it. Of the several different indices presented in
the literature, many of them are based on zonal means of sea level
pressure or geopotential height \citep{ho2012}. \citet{gong1999} defined
the SAM index as the zonal mean sea level pressure differene between
40\degree S and 60\degree S, which is also the definition used by the
station-based index in \citet{marshall2003}. \citet{baldwin2009}
proposed defining the Northern and Annular modes as the leading EOF of
the zonally averaged geopotential height at each level.

Even though these indices are based on zonal averages, the spatial the
spatial structure of the SAM computed from them contains noticeable
deviations from zonal symmetry, particularly in the Pacific Ocean
region. These zonal asymmetries have not been widely studied, but
previous work suggest that they strongly modulate the regional impacts
of the SAM, going as far as reversing its relationship between
precipitation in South America \citep{silvestri2009}. At the very least,
the fact that the SAM is not entirely zonally symmetric hinders our
ability to reconstruct its historical variability prior to the
availabilty of dense observations in the Southern Hemisphere
\citep{jones2009}.

At least some of the variabilty associated with the zonal asymmetries of
the SAM is probably forced by the tropics. In particular, ENSO-like
variability affects the Southern Hemisphere extratopics through the
Pacific-South American Pattern \citep{mo1987, kidson1988, karoly1989},
whose wave train proyects strongly onto the zonal anomalies
corresponding to the SAM in the Pacific sector. And although the
relationship between ENSO and SAM is far from simple, tropical
influences on the SAM have been observed
\citep{fan2007, fogt2011, clem2013}.

Positive trends in SAM index have been documented by various researchers
using different indices mostly on boreal Summer and Autumn \citep[ and
references therein]{fogt2020}. These trends are thought of driven
primarily by stratospheric ozone depletion and understood in the context
of zonal mean variables. However, it's not clear how or if the
asymmetric component responds to this forcing, or whether its variabilty
could be masking influencing the observed trends independendly.

Similarly unclear are the specific impacts of the zonally asymmetric
component of the SAM. Positive phase of the SAM is associated with
generally colder temperatures over Antarctica and warmer temperatures at
higher latitudes \citep{jones2019} (and viceversa for negative SAM), but
there are significant deviations from this zonal mean response, notably
in the Antarctic Peninsula and the South Atlantic \citep{fogt2012}. The
SAM signal in precipitation behaves similarly, although with even
greater deviation from zonal symmetry \citep{lim2016}. The importance of
zonal asymmetries of the SAM in these impacts have been studied in
certain regions. For example, the SAM-precipitation relationship in
Southeastern South America and Southern Brazil can be explained by the
PSA-like zonally asymetric circulation associated with the SAM
\citep{silvestry2009, rosso2018}. \citet{fan2007} also found that
precipitation in East Asia was impacted by the variabilty of only the
Western Hemisphere part of the SAM.

We are not aware of any previous work which quantifies the temporal
variability of the asymmetric component of the SAM with the exception of
\citet{fogt2012}. However, their methods based on composites of positive
and negative SAM events leads to some issues, such as spatial patterns
derived from as little as 4 cases and from imbalanced periods (for
example, 5 of the 7 cases in their DJF SAM+ composite are from later
than 1988, whereas all of the 8 years in their DJF SAM- composite are
from earlier than 1988). This is particularly important due to the
inhomogeneities in renalysis products prior to the satellite era and the
possible change in the asymmetric structure of the SAM
\citep{silvestri2009}.

Moreover, \citet{fogt2012} studied the zonal asymmetric component of the
SAM only in sea level pressure. Zonal asymmetries in the SAM spatial
pattern are fairly barotropic throughout the troposphere, but they
change dramatically in the stratosphere \citep{baldwin2009}.

Our objective is, then, to systematically characterise the zonally
asymmetric and symmetric components of the SAM variability. For each
level, we construct two indices which aim to capture exclusively the
variability of the symmetric and asymmetric component respectively. We
assess their vertical structure and coherence, temporal variability and
trends. We then study the spatial patterns described by the variability
exclusive to each index. Finally, we investigate their relationship with
temperature and precipitation anomalies.

In the Section \ref{sec:methods} we describe the methods. In Section
\ref{sec:results}.\ref{sec:temporal} we describe the temporal
variability and vertical coherence of the inidices. In Section
\ref{sec:results}.\ref{sec:spatial}, we analyse the spatial patterns of
geopotential height associated with them. In Section
\ref{sec:results}.\ref{sec:impacts}, we study their relationship with
surface-level tempreature and precipitation.

@@

\label{sec:methods}

@@

To describe the Southern Annular Mode and its variability we used
monthly geopotential height at 2.5\degree~longitude by
2.5\degree~latitude of horizontal resolution and 37 vertical isobaric
levels from ERA5 \citep{hersbach2020} for the period 1979 to 2018. We
restrict our analysis to the post-satellite era to avoid any confounding
factors arising from the introduction of satellite observations.

We describe the relationship between the SAM indices and temperature and
precipitation. We use temperature data from NOAA's Merged Land Ocean
Global Surface Temperature Analysis V4.0.1 \citep{vose2012, smith2008},
which blends land and ocean temperature analysis into a monthly global
grid 5\degree~ longitude by 5\degree~ latitude For precipitation, we use
monthly, 0.5\degree~ longitude by 0.5\degree~ latitude data from the
Global Precipitation Climatology Centre \citep{schneider2015}.

@@

\label{sec:results}

@@

\label{sec:temporal}

@@

\label{sec:spatial}

@@

\label{sec:impacts}

@@

The anomalous circulation at 700 hPa associated with the Symmetric SAM
(panel c.1) indicate anomalous Easterly flow over Southern Chile. This
leads to reduced influx of moist air from the Pacific Ocean which, is
the main source of precipitable water in that region. On the other hand,
the anomalous circulation associated with positive values of Asymmetric
SAM (panel b.1) in the Atlantic is anticyclonic in the South and
cyclonic in the North. This creates anomalous South-Easterly flow over
Southeastern South America, which inhibits the flow of the Low Level Jet
to the region \citep{silvestri2009, zamboni2010}. This same pattern was
found to be associated with increased precipitation in Southern Brazil
during South Atlantic Convergence Zone events \citep{rosso2018}.

@@

The two-year periodcty we found in the stratospheric Symmetric SAM might
point to a link between the SAM and the Quasi Biennial Oscillation.
There is evidence of influence between the QBO and the Northern Annular
Mode \citep[e.g.][]{holton1980, watson2014, zhang2020}, so it's not
unlikely that the SAM would be similarly connected. However establishing
this link would require further research.
